HMM

### 2015-12-30 paper3.Rnw
need to do Ben's notes one by one, Makefile, make sure to gitroot

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
\todo{do we want to call these transition probabilities (throughout)?}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\todo{what do we think of this subsection title?}

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets. \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

>>>>

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
\todo{do we want to call these transition probabilities (throughout)? ML: I prefer to call it this because it is consistent with explaning HMM parts.}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\todo{what do we think of this subsection title? ML:sounds good to me}

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets. \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats? ML: Yes, I have all the plots ready in the figures folder. (Note, the public catID is different from what Madelon used, so we have to be careful)}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work. 
ML: conclusion is good, I still need to add viterbi paragraph}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever) ML: What do you think? Are ecologist more comfortable with modes/ latent states?}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.
